Perturbation theory in the complex plane: Exceptional points and where to find them

### 2020-11-14 Manuscript/EPAWTFT.tex
saving work on Hubbard

\newcommand{\Ne}{N}
\newcommand{\hI}{\Hat{I}}
\newcommand{\hH}{\Hat{H}}
\newcommand{\hS}{\Hat{S}}
\newcommand{\hT}{\Hat{T}}
\newcommand{\hW}{\Hat{W}}
\newcommand{\hV}{\Hat{V}}
\newcommand{\hc}[2]{\Hat{c}_{#1}^{#2}}
\newcommand{\hn}[1]{\Hat{n}_{#1}}
\newcommand{\n}[1]{n_{#1}}
\newcommand{\Dv}{\Delta v}
\newcommand{\up}{\uparrow}
\newcommand{\dw}{\downarrow}
\newcommand{\updw}{\up\dw}

\newcommand{\ra}{\rightarrow}
\newcommand{\pt}{$\mathcal{PT}$}

@@

%===================================%
\subsection{An illustrative example}
%===================================%
In order to highlight the general properties of EPs mentioned above, we propose to consider the following $2 \times 2$ Hamiltonian commonly used in quantum chemistry

\begin{equation}
\label{eq:H_2x2}
	\bH = 
	\begin{pmatrix}
		\epsilon_1	&	\lambda	\\
		\lambda		&	\epsilon_2
	\end{pmatrix},
\end{equation}
which represents two states of energies $\epsilon_1$ and $\epsilon_2$ coupled with a strength of magnitude $\lambda$.
This Hamiltonian could represent, for example, a minimal-basis configuration interaction with doubles (CID) for the hydrogen molecule. \cite{SzaboBook}

For real $\lambda$, the Hamiltonian \eqref{eq:H_2x2} is clearly Hermitian, and it becomes non-Hermitian for any complex $\lambda$ value.
Its eigenvalues are
\begin{equation}
\label{eq:E_2x2}
	E_{\pm} = \frac{\epsilon_1 + \epsilon_2}{2} \pm \frac{1}{2} \sqrt{(\epsilon_1 - \epsilon_2)^2 + 4\lambda^2}.
\end{equation}
%and they are represented as a function of $\lambda$ in Fig.~\ref{fig:2x2}.
One notices that the two states become degenerate only for a pair of complex conjugate values of $\lambda$
\begin{equation}
\label{eq:lambda_EP}
	\lambda_\text{EP} = \pm i\,\frac{\epsilon_1 - \epsilon_2}{2},
\end{equation}
with energy
\begin{equation}
\label{eq:E_EP}
	E_\text{EP} = \frac{\epsilon_1 + \epsilon_2}{2},
\end{equation}
which correspond to square-root singularities in the complex-$\lambda$ plane. % (see Fig.~\eqref{fig:2x2}).
These two $\lambda$ values, given by Eq.~\eqref{eq:lambda_EP}, are the so-called EPs and one can clearly see that they connect the ground and excited states.
Starting from $\lambda_\text{EP}$, two square-root branch cuts run on the imaginary axis towards $\pm i \infty$.
In the real $\lambda$ axis, the point for which the states are the closest ($\lambda = 0$) is called an avoided crossing and this occurs at $\lambda = \Re(\lambda_\text{EP})$.
The ``shape'' of the avoided crossing is linked to the magnitude of $\Im(\lambda_\text{EP})$: the smaller $\Im(\lambda_\text{EP})$, the sharper the avoided crossing is.

Around $\lambda = \lambda_\text{EP}$, Eq.~\eqref{eq:E_2x2} behaves as \cite{MoiseyevBook}
\begin{equation} %\label{eq:E_EP}
        E_{\pm} = E_\text{EP} \pm \sqrt{2\lambda_\text{EP}} \sqrt{\lambda - \lambda_\text{EP}},
\end{equation}
and following a complex contour around the EP, \ie, $\lambda = \lambda_\text{EP} + R \exp(i\theta)$, yields
\begin{equation}
        E_{\pm}(\theta) = E_\text{EP} \pm \sqrt{2\lambda_\text{EP} R}  \exp(i\theta/2),
\end{equation}
and we have
\begin{align}
	E_{\pm}(2\pi) & = E_{\mp}(0),
	&
	E_{\pm}(4\pi) & = E_{\pm}(0). \notag
\end{align}
This evidences that encircling non-Hermitian degeneracies at EPs leads to an interconversion of states, and two loops around the EP are necessary to recover the initial energy.
Additionally, the wave function picks up a geometric phase and four loops are required to recover the starting wave function.
\cite{MoiseyevBook}

@@

\subsection{M{\o}ller-Plesset perturbation theory}

>>>>

\newcommand{\Ne}{N}
\newcommand{\hI}{\Hat{I}}
\newcommand{\hH}{\Hat{H}}
\newcommand{\hS}{\Hat{S}}
\newcommand{\hT}{\Hat{T}}
\newcommand{\hW}{\Hat{W}}
\newcommand{\hV}{\Hat{V}}
\newcommand{\hc}[2]{\Hat{c}_{#1}^{#2}}
\newcommand{\hn}[1]{\Hat{n}_{#1}}
\newcommand{\n}[1]{n_{#1}}
\newcommand{\Dv}{\Delta v}

\newcommand{\ra}{\rightarrow}
\newcommand{\up}{\uparrow}
\newcommand{\dw}{\downarrow}

\newcommand{\updot}{%
  \mathrel{\ooalign{\hfil$\vcenter{
    \hbox{$\scriptscriptstyle\bullet$}}$\hfil\cr$\uparrow$\cr}
  }%
}
\newcommand{\dwdot}{%
  \mathrel{\ooalign{\hfil$\vcenter{
    \hbox{$\scriptscriptstyle\bullet$}}$\hfil\cr$\downarrow$\cr}
  }%
}
\newcommand{\vac}{%
  \mathrel{\ooalign{\hfil$\vcenter{
    \hbox{$\scriptscriptstyle\bullet$}}$\hfil\cr$ $\cr}
  }%
}

\newcommand{\uddot}{%
  \mathrel{\ooalign{\hfil$\vcenter{
    \hbox{$\scriptscriptstyle\bullet$}}$\hfil\cr$\uparrow\downarrow$\cr}
  }%
}

@@

%===================================%
\subsection{An illustrative example}
%===================================%
In order to highlight the general properties of EPs mentioned above, we propose to consider the ubiquitous symmetric Hubbard dimer at half filling (\ie, with two opposite-spin fermions) whose Hamiltonian reads in the singlet configuration state function basis 
%\begin{align}
%		\ket{1\up1\dw}		&	\ket{1\up2\dw}		&	\ket{1\dw2\up}		&	\ket{2\up2\dw}	\\
%		\uddot \quad \vac	&	\updot \quad \dwdot	&	\dwdot \quad \updot	&	\vac \quad \uddot	 \\
%\end{align}
\begin{equation}
\label{eq:H_FCI}
	\bH = 
	\begin{pmatrix}
		-2t	+ U &	0 & U/2	\\
		0		&	U 	&	0	\\
		U/2 &	0 & -2t	+ U	\\
	\end{pmatrix},
\end{equation}
where $t$ is the hopping parameter and $U$ is the on-site Coulomb repulsion.
We refer the interested reader to Refs.~\onlinecite{Carrascal_2015,Carrascal_2018} for more details about this system.
We will consistently use this system to illustrate the different concepts discussed in the present review article.

For real $U$, the Hamiltonian \eqref{eq:H_FCI} is clearly Hermitian, and it becomes non-Hermitian for any complex $U$ value.
The eigenvalues associated with its singlet ground state and singlet doubly-excited state are
\begin{equation}
\label{eq:E_FCI}
	E_{\pm} = \frac{1}{2} \qty( U \pm \sqrt{(4t^2) + U^2} ).
\end{equation}
and they are represented as a function of $U$ in Fig.~\ref{fig:FCI} together with the energy of the singlet open-shell configuration of energy $U$.

%%% FIG 1 %%%
\begin{figure*}
	\includegraphics[height=0.3\textwidth]{fig1a}
	\hspace{0.1\textwidth}
	\includegraphics[height=0.3\textwidth]{fig1b}	
	\caption{
	Exact energies for the Hubbard dimer as functions of $U/t$.
	\label{fig:FCI}}
\end{figure*}

One notices that these two states become degenerate only for a pair of complex conjugate values of $U$
\begin{equation}
\label{eq:lambda_EP}
	U_\text{EP} = \pm 4 i t,
\end{equation}
with energy
\begin{equation}
\label{eq:E_EP}
	E_\text{EP} = \pm 2 i t,
\end{equation}
which correspond to square-root singularities in the complex-$U$ plane [see Fig.~\eqref{fig:FCI}].
These two $U$ values, given by Eq.~\eqref{eq:lambda_EP}, are the so-called EPs and one can clearly see that they connect the ground and excited states.
Starting from $U_\text{EP}$, two square-root branch cuts run on the imaginary axis towards $\pm i \infty$.
In the real $U$ axis, the point for which the states are the closest ($U = 0$) is called an avoided crossing and this occurs at $U = \Re(U_\text{EP})$.
The ``shape'' of the avoided crossing is linked to the magnitude of $\Im(U_\text{EP})$: the smaller $\Im(U_\text{EP})$, the sharper the avoided crossing is.

Around $U = U_\text{EP}$, Eq.~\eqref{eq:E_FCI} behaves as \cite{MoiseyevBook}
\begin{equation} %\label{eq:E_EP}
        E_{\pm} = E_\text{EP} \pm \sqrt{2U_\text{EP}} \sqrt{U - U_\text{EP}},
\end{equation}
and following a complex contour around the EP, \ie, $U = U_\text{EP} + R \exp(i\theta)$, yields
\begin{equation}
        E_{\pm}(\theta) = E_\text{EP} \pm \sqrt{2U_\text{EP} R}  \exp(i\theta/2),
\end{equation}
and we have
\begin{align}
	E_{\pm}(2\pi) & = E_{\mp}(0),
	&
	E_{\pm}(4\pi) & = E_{\pm}(0). \notag
\end{align}
This evidences that encircling non-Hermitian degeneracies at EPs leads to an interconversion of states, and two loops around the EP are necessary to recover the initial energy.
Additionally, the wave function picks up a geometric phase and four loops are required to recover the starting wave function.
\cite{MoiseyevBook}

@@

Coming back to the Hubbard dimer, the HF energy is
\begin{equation}
	E_\text{HF} = -t \qty[ \sin \theta_\alpha  + \sin \theta_\beta ] + \frac{U}{2} \qty[ 1 + \cos \theta_\alpha \cos \theta_\beta ]
\end{equation}
where 
\begin{align}
	\psi_{1\sigma} & = \cos(\frac{\theta_\sigma}{2}) s_1 - \sin(\frac{\theta_\sigma}{2})s_2
	\\
	\psi_{2\sigma} & = \sin(\frac{\theta_\sigma}{2}) s_1 + \cos(\frac{\theta_\sigma}{2})s_2 
\end{align}

\begin{align}
	\theta_\alpha & = \arctan (-\frac{\sqrt{U^2 - 4t^2}}{U},\frac{2t}{U}) 
	\\
	\theta_\beta & = \arctan (+\frac{\sqrt{U^2 - 4t^2}}{U},\frac{2t}{U}) 
\end{align}

%=====================================================%
\subsection{M{\o}ller-Plesset perturbation theory}
%=====================================================%
